---
title: "library(ggplot2)"
author: "Kyle Chung"
date: "July 5, 2015"
output: 
    ioslides_presentation:
      keep_md: yes
      smaller: yes
      transition: faster
      widescreen: yes
---

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
```

# basic syntax

## the use of main function: *ggplot*

+ `library(ggplot2)`
+ `ggplot(data, aes(x, y, group, ...)) + geom_obj(...) + modify()`
+ or a quick and dirty alternative: `qplot`
+ Usage:
    + data: an object of class `data.frame`
        + **DATA** of your plot
    + aes: a function that returns aesthetic mappings
        + **VARIABLE** in data to be plotted
    + geom_obj: geometric objects
        + **TYPE** of plot
        + `geom_bar()`, `geom_line()`, `geom_point()`, ...
    + modify: additional modification to the plot
        + `ggtitle()`, `scale_x_discrete()`, `theme_minimal()`...


## a quick example | plot distribution of diamond cut quality

```{r, fig.retina=2}
ggplot(data=diamonds, aes(x=cut)) + geom_bar()
```


## Factor or Numeric?
+ Variable class affects `ggplot`'s behavior
+ Variable class affects `ggplot`'s behavior. **Twice**.
+ Always check your data.frame (use `str` or `class`) before calling `ggplot`


# bar plot

## sample dataset: `diamonds`

```{r}
str(diamonds)
head(diamonds)
```


## x must be of type `factor`

```{r, fig.retina=2}
ggplot(data=diamonds, aes(x=cut)) + geom_bar()
```


## geom_bar drops category with no occurrence

```{r, fig.retina=2}
ggp <- ggplot(data=diamonds[1:5,], aes(x=cut)) + geom_bar()
ggp
```


## force display all categories

```{r, fig.retina=2}
ggp <- ggplot(data=diamonds[1:5,], aes(x=cut)) + geom_bar()
### <b>
ggp + scale_x_discrete(drop=FALSE) # ?scale_x_discrete
### </b>
```


## horizontal bar

```{r, fig.retina=2}
ggplot(data=diamonds, aes(x=cut)) + geom_bar() + coord_flip()
```


## change label naming

```{r, fig.retina=2}
ggp <- ggp + xlab("Cut") + ylab("Count") + ggtitle("Hello ggplot!")
ggp
```


## change color

```{r, fig.retina=2}
# want to customize colors? refer to: www.cookbook-r.com/Graphs/Colors_(ggplot2)/
ggp + geom_bar(fill="snow", color="black") # see colors() if you're picky
```


## plot counts as is | when counts are pre-calculated

```{r}
diamonds_precounted <- as.data.frame(table(diamonds$cut, dnn=c("Cut")))
diamonds_precounted
```
+ take a guess before you try!
+ `ggplot(data=diamonds_precounted, aes(x=Cut)) + geom_bar() # epic FAIL`


## | wrong!

```{r, fig.retina=2}
ggplot(data=diamonds_precounted, aes(x=Cut)) + geom_bar()
```


## | correct!

```{r, fig.retina=2}
ggplot(diamonds_precounted, aes(x=Cut, y=Freq)) + geom_bar(stat="identity") # default is "bin"
```


## when using `stat="identity"`

+ row should be unique
    + otherwise counts will be summed up
+ missing label will be present at default
    + differ from `stat="bin"`
+ negative bar is allowed
```{r}
diamonds[1:5,]
```


## 

```{r, fig.retina=2}
ggplot(diamonds[1:5,], aes(x=cut, y=depth)) + geom_bar(stat="identity")
```


## reorder x

```{r, fig.retina=2}
ggplot(diamonds_precounted, aes(x=reorder(Cut, -Freq), y=Freq)) + 
  geom_bar(stat='identity') # The order is determined by factor levels
```


## stack grouping | by `fill`

```{r, fig.retina=2}
ggplot(data=diamonds, aes(x=color, fill=cut)) + geom_bar()
```


## stack grouping |  by `color`

```{r, fig.retina=2}
ggplot(data=diamonds, aes(x=color, color=cut)) + geom_bar()
```


## dodge grouping

```{r, fig.retina=2}
ggplot(data=diamonds, aes(x=color, fill=cut)) + geom_bar(position="dodge")
```


## from bar to histogram | when x is numeric

```{r, fig.retina=2, message=FALSE}
ggplot(data=diamonds, aes(x=price)) + geom_bar()
```


## histogram grouping | stack

```{r, fig.retina=2, message=FALSE}
ggplot(data=diamonds, aes(x=price, fill=cut)) + geom_bar(position="stack")
```


## histogram grouping | identity (overlapping)

```{r, fig.retina=2, message=FALSE}
ggplot(data=diamonds, aes(x=price, fill=cut)) + geom_bar(position="identity", alpha=.5)
```


## from histogram to density plot

```{r, fig.retina=2}
ggplot(data=diamonds, aes(x=price, fill=cut)) + geom_density(position="identity", alpha=.5)
```


## both histogram and density (scale wrong...)

```{r, fig.retina=2, message=FALSE}
ggplot(data=diamonds[diamonds$cut %in% c("Fair", "Ideal"),], aes(x=price, fill=cut)) + 
    geom_density(position="identity", alpha=.5) + 
    geom_bar(position="identity", alpha=.5)
```


## both histogram and density

```{r, fig.retina=2, message=FALSE}
ggplot(data=diamonds[diamonds$cut %in% c("Fair", "Ideal"),], aes(x=price, y=..density.., fill=cut)) + 
    geom_density(position="identity", alpha=.5) + 
    geom_bar(position="identity", alpha=.5)
```


## exercise!

+ use the weather data to plot:
    + total rainy hours for each city (PrecpHour)
    + total rainy volume for each city (Precp)
    + total sunny hours for each city (SunShine)
    + total hours grouped by rainy / sunny for each city
+ remember to check variable type in advanced
+ the forth plot is rather difficult
```{r, echo=FALSE}
src_path <- "/Users/kylechung/Dropbox/DSC2015/ggplot_lecture/data"
citynames <- unique(gsub("_.*", '', dir(src_path)))
citynames <- citynames[1:4] # take english version only
dat <- list()
for ( city in citynames ) {
    dat[[city]] <- do.call(rbind, lapply(grep(city, dir(src_path, full.names=TRUE), value=TRUE), read.csv, stringsAsFactors=FALSE))
    dat[[city]]$city <- city
}
dat <- do.call(rbind, dat)

dat[dat$Precp == 'T', "Precp"] <- '0'
dat$Precp <- as.numeric(dat$Precp)
```


## exercise! | total rainy hours by city

```{r, echo=FALSE}
ggplot(dat, aes(x=city, y=PrecpHour)) + geom_bar(stat="identity")
```


## exercise! | total rainy volume by city

```{r, echo=FALSE}
ggplot(dat, aes(x=city, y=Precp)) + geom_bar(stat="identity")
```


## exercise! | total sunny hours by city

```{r, echo=FALSE}
ggplot(dat, aes(x=city, y=SunShine)) + geom_bar(stat="identity")
```


## exercise! | hours group by rainy/suny by city

```{r, echo=FALSE}
dd <- by(dat, dat$city, 
         function(x) c(rainyHour=sum(x$PrecpHour), sunnyHour=sum(x$SunShine)))
dd <- as.data.frame(do.call(rbind, dd))
dd$city <- rownames(dd)
dd <- reshape(dd, direction="long", varying=c("rainyHour", "sunnyHour"), 
              idvar="city", v.names="hour", times=c("rainyHour", "sunnyHour"), timevar="status")
ggplot(dd, aes(x=city, y=hour, fill=status)) + geom_bar(stat="identity")
```


## exercise: possible solution 1 (basic)

```{r, eval=FALSE}
librart(ggplot2)
src_path <- "/Users/kylechung/Dropbox/DSC2015/ggplot_lecture/data"
citynames <- unique(gsub("_.*", '', dir(src_path)))
citynames <- citynames[1:4] # take english version only
dat <- list()
for ( city in citynames ) {
    dat[[city]] <- do.call(rbind, lapply(grep(city, dir(src_path, full.names=TRUE), value=TRUE), 
                                         read.csv, stringsAsFactors=FALSE))
    dat[[city]]$city <- city
}
dat <- do.call(rbind, dat)
dat[dat$Precp == 'T', "Precp"] <- '0'
dat$Precp <- as.numeric(dat$Precp)

ggplot(dat, aes(x=city, y=PrecpHour)) + geom_bar(stat="identity")
ggplot(dat, aes(x=city, y=Precp)) + geom_bar(stat="identity")
ggplot(dat, aes(x=city, y=SunShine)) + geom_bar(stat="identity")

dd <- by(dat, dat$city, 
         function(x) c(rainyHour=sum(x$PrecpHour), sunnyHour=sum(x$SunShine)))
dd <- as.data.frame(do.call(rbind, dd))
dd$city <- rownames(dd)
dd <- reshape(dd, direction="long", varying=c("rainyHour", "sunnyHour"), 
              idvar="city", v.names="hour", times=c("rainyHour", "sunnyHour"), timevar="status")
ggplot(dd, aes(x=city, y=hour, fill=status)) + geom_bar(stat="identity")
```


## exercise: possible solution 2 (advanced)

```{r, eval=FALSE}
library(ggplot2)
library(data.table) # need 1.9.5 => devtools:install_github("Rdatatable/data.table", build_vignettes=FALSE)
src_path <- "/Users/kylechung/Dropbox/DSC2015/ggplot_lecture/data"
citynames <- unique(gsub("_.*", '', dir(src_path)))
citynames <- citynames[1:4] # take english version only
dat <- list()
for ( city in citynames )
    dat[[city]] <- rbindlist(lapply(grep(city, dir(src_path, full.names=TRUE), value=TRUE), fread))
dat <- rbindlist(dat, idcol="city")
dat[Precp == 'T', Precp:='0']
dat[, Precp:=as.numeric(Precp)]

for ( var in c("PrecpHour", "Precp", "SunShine") ) 
    print(ggplot(dat, aes_string(x="city", y=var)) + geom_bar(stat="identity"))

dd <- melt(dat[, list(PrecpHour=sum(PrecpHour), SunShine=sum(SunShine)), by="city"], 
           id.vars="city", measure.vars=c("PrecpHour", "SunShine"), 
           variable.name="status", value.name="hour")
ggplot(dd, aes(x=city, y=hour, fill=status)) + geom_bar(stat="identity")
```


<div class="columns-2">
## proportional stacking bar

```{r, fig.retina=2, fig.width=5, fig.height=5.5, echo=FALSE}
ggplot(data=diamonds, aes(x=color, fill=cut)) + geom_bar() + theme_minimal()

diamonds$Count <- 0L
diamondsDF <- aggregate(data=diamonds, Count ~ color + cut, FUN=length)
diamondsDF <- do.call(rbind, lapply(split(diamondsDF, diamondsDF$color), 
                                  function(x) within(x, pct <- Count/sum(Count))))
ggplot(diamondsDF, aes(x=color, y=pct, fill=cut)) + geom_bar(stat='identity') + theme_minimal()
```
</div>


<div class="columns-2">
## need to transform data and then use `stat="identity"`

```{r, echo=FALSE}
library(data.table)
diamondsDT <- as.data.table(diamonds)
diamondsDT <- diamondsDT[, list(cnt=.N), by="color,cut"]
diamondsDT <- diamondsDT[, pct:=cnt / sum(cnt), by="color"]
head(diamondsDT, 20)
```

```{r, echo=FALSE}
table(diamonds$color, diamonds$cut)
```
</div>


## possible solution 1 | just do it: `aggregate` and `lapply`

```{r}
diamonds$cnt <- 0L
diamondsDF <- aggregate(data=diamonds, cnt ~ color + cut, FUN=length)
diamondsDF <- do.call(rbind, lapply(split(diamondsDF, diamondsDF$color), 
                                  function(x) within(x, pct <- cnt / sum(cnt))))
head(diamondsDF)
```


## possible solution 2 | play tricky: `table` and `by`

```{r}
DF <- as.data.frame(table(diamonds$color, diamonds$cut, dnn=list("color", "cut")))
DF <- do.call(rbind, by(DF, DF$color, 
                        function(x) within(x, pct <- Freq / sum(Freq))))
head(DF)
```


## possible solution 3 | try `ave`
```{r}
diamonds$cnt <- ave(rep(1, nrow(diamonds)), diamonds$color, diamonds$cut, FUN=sum)
diamonds$cnt_by_color <- ave(rep(1, nrow(diamonds)), diamonds$color, FUN=sum)
diamonds <- transform(diamonds, pct=cnt / cnt_by_color)
DFave <- unique(diamonds[,c("color", "cut", "cnt", "pct")])
head(DFave)
```


## possible solution 4 | use `data.table`

```{r}
library(data.table)
diamondsDT <- as.data.table(diamonds)
diamondsDT <- diamondsDT[, list(cnt=.N), by="color,cut"]
diamondsDT <- diamondsDT[, pct:=cnt / sum(cnt), by="color"]
head(diamondsDT)
```


## possible solution 5 | use `dplyr`

```{r, message=FALSE}
library(dplyr)
DP <- summarise(group_by(diamonds, color, cut), cnt=n())
DP <- mutate(DP, pct=cnt / sum(cnt))
head(DP)
```


## possible solution 6 | forever SQL

```{r, message=FALSE}
# options(gsubfn.engine='R') # try this line should you have trouble loading the package
library(sqldf)
tmp1 <- sqldf('select color, cut, count(*) as cnt from diamonds group by color, cut')
tmp2 <- sqldf('select color, sum(cnt) as cnt_by_color from tmp1 group by color')
DFSQL <- sqldf('select tmp1.color, cut, cnt, (cnt*1.0 / cnt_by_color) as pct 
               from tmp1 join tmp2 on tmp1.color = tmp2.color')
head(DFSQL)
```


## finally, proportional stacking bar

```{r, fig.retina=2, echo=FALSE}
ggplot(diamondsDF, aes(x=color, y=pct, fill=cut)) + geom_bar(stat='identity')
```


<div class="columns-2">
## exercise!

```{r, fig.retina=2, fig.width=5, fig.height=5.5, echo=FALSE}
ggplot(dd, aes(x=city, y=hour, fill=status)) + geom_bar(stat="identity")
dd2 <- do.call(rbind, by(dd, dd$city, 
                        function(x) within(x, pct <- hour / sum(hour))))
ggplot(dd2, aes(x=city, y=pct, fill=status)) + geom_bar(stat="identity")
```
</div>


## possible solution
```{r, eval=FALSE}
library(ggplot2)
dd2 <- do.call(rbind, by(dd, dd$city, 
                        function(x) within(x, pct <- hour / sum(hour))))
ggplot(dd2, aes(x=city, y=pct, fill=status)) + geom_bar(stat="identity")
```


